---
title: "ESM 206 Lab 5"
author: "Minnie Ringland"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***
#### Lab 5 Objectives

- Learn to parse dates with 'lubridate'
- Get counts of observations with count()
- Uncount() rows
- One and two-sample t-test
- Create a heatmap with geom_tile()

```{r, include =  FALSE}
# include = FALSE means that code will run but output will not show up in the html

# Now attach packages and read in data

library(tidyverse)
library(here)
library(janitor)

lobster_abundance <- read_csv(here::here("data", "lobster_abundance.csv"),
                              na = "-99999") %>% 
  clean_names()

```

***
Use tidyr::uncount() function to convert our lobster data from frequency format to case format

```{r}
lobster_tidy <- lobster_abundance %>%
  tidyr::uncount(lobster_count)
```


***
#### Exploratory data visualization

We're only going to consider 'site' as our variable of interest.

```{r, echo = FALSE , warning = FALSE , message = FALSE }

ggplot(lobster_tidy, aes(x = site, y = size_mm)) +
  geom_jitter(aes(color = site),
              alpha = 0.5,
              width = 0.2)


# Try a histogram
# scales in facet_wrap allow graphs to vary in scale
ggplot(lobster_tidy, aes(x = size_mm)) +
  geom_histogram(aes(fill = site)) +
  facet_wrap(~site, scales = "free")


# Now we'll look at a QQ plot to see if the data follows a normal distribution
ggplot(lobster_tidy, aes(sample = size_mm)) +
  geom_qq() +
  facet_wrap(~site)


```

***
#### Convert the 'date' column to class 'Date'  

We'll use the 'lubridate' package to convert to Date format, and then help us eaily parse month & year

```{r}

lobster_date <- lobster_tidy %>% 
  mutate(
    date_new = lubridate::mdy(date)
  )

# class(lobster_date$date_new)

```

Now let's parse year and month using lubridate::month() and lubridate::year()

```{r}

lobster_parse_date <- lobster_date %>% 
  mutate(
    obs_month = lubridate::month(date_new, label = TRUE),
    obs_year = lubridate::year(date_new)
  )

# class(lobster_parse_date$obs_month)
# This is now an ordered factor - saves us a lot of work

```

Now let's find counts of observed lobsters based on different hierarchical groupings.  
We will be using dplyr count().

First, let's count lobsters by year and month:

```{r}

lobster_ym <- lobster_parse_date %>% 
  dplyr::count(obs_year, obs_month)

lobster_ym

lobster_y <- lobster_parse_date %>% 
  dplyr::count(obs_year)

lobster_y

lobster_site <- lobster_parse_date %>% 
  dplyr::count(site)

lobster_site

```

If we want to create a summary table that contains statistics OTHER than counts by group, it's easier to use group_by() + n()

```{r}
lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE),
    lobster_number = n()
  )

lobster_summary
```

***
#### Find confidence intervals

Use t.test() function from base R package to find confidence intervals (for one sample) and perform t-tests to compare means of two samples (...this will be covered conceptually in lectures Week 6)

```{r}

ivee_lobster <- lobster_tidy %>% 
  filter(site == "IVEE") %>% 
  pull(size_mm) 

# Can use pull() to grab a column of values and turn it into a vector

t.test(ivee_lobster)

```


***
#### Two-sample t-test to compare means
(are these samples different enough to be considered separate populations?)

We want to ask: Is there a significant different in lobster lengths at Naples and Mohawk reefs?

We've done our necessary exploratory analyses to determine that a two-sample t-test for means comparison is appropriate.

```{r}
# Two methods

# First: pull a vector

napl_sample <- lobster_tidy %>% 
  filter(site == "NAPL") %>% 
  pull(size_mm)

mohk_sample <- lobster_tidy %>% 
  filter(site == "MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(napl_sample, mohk_sample)

mn_ttest


```

We summarize this result by saying there is a significant difference in lobster lengths between Naples and Mohawk Reefs (t(df) = statistic, p < 0.001, alpha - 0.05).

```{r}
# Second method of model

lobster_mn <- lobster_tidy %>% 
  filter(site %in% c("NAPL","MOHK"))

mn_ttest2 <- t.test(size_mm ~ site, data = lobster_mn)

mn_ttest2


```

***
How to do in-line summary? (Rmd)
Can use t.test() values

```{r}
mn_ttest2$p.value
mn_ttest2$statistic

```

Here is the p-value: `r mn_ttest2$p.value` 
Here is the t-statistic: `r mn_ttest2$statistic`


***
#### Now a heatmap!

```{r}

lobster_ys <- lobster_parse_date %>%
  count(obs_year, site)

ggplot(data = lobster_ys, aes(x = obs_year, y = site)) +
  geom_tile(aes(fill = n))


```

